A hazards approach to the biometric analysis of infant mortality

### 2023-03-06 paper.Rmd
first commit

```{r logcube, fig.cap = "The log-cube transform of time in days versus the identity transform (dashed). Note that g(365) = 365.", fig.height = 7}
x <- seq(0, 365, length = 366)
plot(x, g(x), type = "l", xlab = "Days", ylab = "g(Days)", 
     col = "blue", axes = FALSE)
att <- c(0, 30, 183, 365)
axis(1, at = att, labels = att)
axis(2, las = 1, at = g(att), labels = c("0", "g(30)", "g(183)", "365"))
box()
lines(c(0, 365), c(0, 365), lty = 2)
abline(v = c(0, 365), h = c(0, 365))
abline(v = 30, h = g(30), col = "darkgreen", lty = 3)
```

@@

Assume that a cohort of infants is followed over time from birth to age one. There are no drop-outs 
(no right censoring). The exact age in days at each observed death is noted and transformed by $g$, 
and due to imperfect time measuring there may be tied death ages. As an illustrative example, 
we use a data set from northern Sweden, covering the years 1861--1950.
```{r readdata}
infant <- readRDS("Rdata/infant.rds")
infant$period <- cut(infant$birthdate, 
                   breaks = as.Date(c("1860-12-31", "1890-12-31", 
                                      "1920-12-31", "1951-01-01")),
                   labels = c("1861-1890", "1891-1920", "1921-1950"))
infant <- infant[, c("birthdate", "sex", "enter", "exit", "event", "period", "icd.chapter")]
ginfant <- infant
ginfant$enter <- g(infant$enter)
ginfant$exit <- g(infant$exit)
```
For the first and last 30 years in this data set, we have the results shown in Figure \@ref(fig:first).

```{r first, fig.cap = "The periods 1861--1890 and 1921--1950, Bourgeois-Pichat method."}
source("R/drawBP.R")
oldpar <- par(mfrow = c(1,2))
##
r1 <- with(ginfant[ginfant$period == "1861-1890", ], eha::Surv(enter, exit, event))
r1 <- eha::risksets(r1, members = FALSE)
drawBP(r1, main = "1861-1890", ylim = c(0, 0.1))
##
##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
##r2 <- eha::risksets(r2, members = FALSE)
##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
##
r3 <- with(ginfant[ginfant$period == "1921-1950", ], eha::Surv(enter, exit, event))
r3 <- eha::risksets(r3, members = FALSE)
drawBP(r3, main = "1921-1950", ylim = c(0, 0.1))
par(oldpar)
```

The cumulative numbers of death are plotted against the
corresponding death ages on the $g$ time scale, and as is seen, the fit to linearity
after 30 days (72 on the $g$ scale) is excellent for both time periods.

@@

Instead of using the total number of births throughout in the denominator oas in  the
Bourgeois-Pichat procedure, we suggest 
using the present riskset size, that is, the the total number of infants still alive 
immediately prior to the death age under investigation. It is easily achieved by the use of 
the function `risksets` in the package `eha`. See Table \@ref(tab:risksets) for 
an illustration.
```{r risksets, echo = FALSE, results = 'asis'}
source("R/tbl.R")
lt1 <- with(r1, cbind(ginv(risktimes), round(risktimes, 1), size, n.events))
lt1 <- data.frame(lt1)
names(lt1) <- c("Day", "g(Day)", "Population", "Deaths")
tbl(lt1[1:7, ], caption = "Life table, 1861-1890.")
```
It starts with `r as.integer(lt1[1, 3])` live births, of which `r lt1[1, 4]` dies on the day of birth, 
on avareage at the age of six hours (a quarter of a day), and so on.

@@

```{r second, fig.cap = "The periods 1861--1890 and 1921--1950, hazard method."}
source("R/drawBB.R")
oldpar <- par(mfrow = c(1,2))
##
##r1 <- with(ginfant[ginfant$period == "1861-1890", ], eha::Surv(enter, exit, event))
##r1 <- eha::risksets(r1, members = FALSE)
drawBB(r1, main = "1861-1890", ylim = c(0, 0.115))
##
##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
##r2 <- eha::risksets(r2, members = FALSE)
##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
##
##r3 <- with(ginfant[ginfant$period == "1921-1950", ], eha::Surv(enter, exit, event))
##r3 <- eha::risksets(r3, members = FALSE)
drawBB(r3, main = "1921-1950", ylim = c(0, 0.115))
par(oldpar)
```

@@

```{r both, fig.cap = "Bourgeois-Pichat and 'hazard method' plots."}
oldpar = par(mfrow = c(1, 2))
plot(r1$risktimes, cumsum(r1$n.events) / r1$size[1], type = "l", col = "blue",
     main = "1861-1890", ylim = c(0, 0.1), xlab = "g(Day)", ylab = "Cum. hazards")
lines(r1$risktimes, cumsum(r1$n.events / r1$size), type = "l", col = "red")
abline(h = 0, v = 0)
text(120, 0.08, "Hazard method", col = "red")
text(200, 0.02,"Bourgeois-Pichat", col = "blue")
##
plot(r3$risktimes, cumsum(r3$n.events) / r3$size[1], type = "l", col = "blue",
     main = "1921-1950", ylim = c(0, 0.1), xlab = "g(Day)", ylab = "Cum. hazards")
lines(r3$risktimes, cumsum(r3$n.events / r3$size), type = "l", col = "red")
abline(h = 0, v = 0)
text(120, 0.05, "Hazard method", col = "red")
text(200, 0.015,"Bourgeois-Pichat", col = "blue")
##
##
par(oldpar)
```

@@

We note that the Bourgeois-Pichat method requires that no right censorings (infants lost 
to follow-up before one year of age) are present, and no left truncations (infants under observation 
only from an age later than birth). The hazards approach, on the other hand, allows
left truncation and right censoring as long as they are *non-informative*. Often, though,
this not very important, since new-born babies tend to be geographically stationary
for their first year of life.

## Post-neonatal mortality on the $g$ scale

The following discussion assumes that age is measured on the $g$ scale. We concentrate on the *postneonatal* period, since that is the period where Bourgeois-Pichat claims that the *cumulative distribution function* (CDF) is *uniform*, and we suggest that the CDF is *exponential*, that is, the *cumulative hazard function* is "uniform" (linear).

In order to see this, the data set from above is *left truncated* at age 30 days (at 72 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 30.

```{r postneo, fig.cap = "Exponential fits to postneonatal data."}
oldpar <- par(mfrow = c(1, 2))
post <- age.window(ginfant, c(g(30), g(365)))
p1 <- post[post$period == "1861-1890", ]
p3 <- post[post$period == "1921-1950", ]
fit1 <- coxreg(Surv(exit - enter, event) ~ 1, data = p1)
wfit1 <- phreg(Surv(exit - enter, event) ~ 1, data = p1, shape = 1)
check.dist(fit1, wfit1, main = "1861-1890")
##
fit3 <- coxreg(Surv(exit - enter, event) ~ 1, data = p3)
wfit3 <- phreg(Surv(exit - enter, event) ~ 1, data = p3, shape = 1)
check.dist(fit3, wfit3, main = "1921-1950")
par(oldpar)
```

The *Exponential* fit is excellent, with a slight edge for the early period data, 
where postneonatal mortality is high. However, one question pops up: If the exponential 
assumption is in fact true, will it also hold for subgroups of the data?

@@

```{r joined, fig.cap = "Exponential fit for two periods."}
p13 <- post[post$period != "1891-1920", ]
p13$period <- factor(p13$period)
fit13 <- coxreg(Surv(exit - enter, event) ~ 1, data = p13)
wfit13 <- phreg(Surv(exit - enter, event) ~ 1, data = p13, shape = 1)
check.dist(fit13, wfit13, main = "1861-1890, 1921-1950")
#
```

@@

## Exogeneous mortality on the $g$ scale

We now consider the whole infant age span (on the $g$ scale), and compare the uniform and exponential densities, especially of interest is the fraction of deaths that occur in the neonatal phase. @manfred04 argues that a less satisfactory property of the *B-P* model is that this fraction is constant, 
19.7 percent, independent of the overall level of infant mortality. This is not the case for the hazards model, see Figure \@ref(fig:exog).
For $p = 0.2$, the fraction in question is 21.5 percent.

@@

## Endogeneous mortality on the $g$ scale

In order to calculate the *endogeneous* infant mortality, a simple three-step procedure leads
to the goal.

@@

2.   Translate the result from 1. to the natural time scale.
We have found that the exogenous mortality distribution on the $g$ scale
has a constant hazard function. Now we want to translate this back to the original, natural time scale.
Apart from obvious reasons, we want to calculate the endogenous mortality by estimating 
the total mortality in the usual way [@nelson72; @aalen78] and subtracting the exogenous mortality, 
all takes of course place on the neonatal part on the natural time scale.
First, it can be established that the relations between the natural and the $g$ time scales
regarding hazard and cumulative hazard functions are given by Equations \@ref(eq:haz) and \@ref(eq:cumhaz).
\begin{align} (\#eq:haz)
h(t) &= \lambda \frac{d}{dt} g(t), \quad 0 \le t \le 365, \\
(\#eq:cumhaz) H(t) &= \lambda g(t), \quad 0 \le t \le 365, 
\end{align}
where $\lambda$ is the constant exogenous hazard on the $g$ scale.
See Figure \@ref(fig:hH).

```{r hH, fig.cap = "Exogenous infant mortality, hazard and cumulative hazard functions."}
lam <- 0.0002 ##rate
oldpar <- par(mfrow = c(1, 2), cex.main = 0.8)
x <- seq(0, 365, length = 1000)
plot(x, lam * dg(x) * 1000, type = "l", main = "Hazard function", col = "blue", 
     axes = FALSE, xlab = "Days", ylab = "per 1000")
axis(1, at = c(0, 30, 183, 365))
axis(2, las = 1)
abline(h = 0, v = 0)
abline(v = 30, lty = 3, col = "darkgreen")
box()
plot(x, lam * g(x) * 1000, type = "l", main = "Cumulative hazard function", col = "blue",
     axes = FALSE, xlab = "Days", ylab = "per 1000")
axis(1, at = c(0, 30, 183, 365))
axis(2, las = 1)
abline(h = 0, v = 0)
abline(v = 30, lty = 3, col = "darkgreen")
box()
par(oldpar)
```

3.   On the full infant interval, estimate the total cumulative hazard rate $A(t)$ with the usual 
Nelson-Aalen estimator $\hat{A}(t)$ [@nelson72; @aalen78]. Then, subtract $H(t)$ from 
$\hat{A}(t)$ to get $\hat{E}(t)$.
\begin{equation}
\hat{E}(t) = \hat{A}(t) - \hat{\lambda} t, 0 < t < g(30).
\end{equation}
Note that subtraction and addition of competing risks are okay on the hazards scale, 
but not with probabilities, which is yet another argument in favor of the hazards approach.

@@

### Step 1: Estimate post-neonatal (exogenous) mortality on the $g$ scale

Post-neonatal mortality is the same as exogenous mortality in the sense that the
hazard functions are the same on the post-neonatal age interval. So the first step
involves data left truncated at age 30 days (or at `r round(g(30))`) on the $g$ scale.
We exemplify the numerical procedure with the aid of the **eha** [@eha; @ehar2].
The first lines of the data set are (Table \@ref(tab:firstlines))
```{r firstlines, results='asis'}

@@

```{r retrunc, echo = TRUE}
postneo <- eha::age.window(infant, c(30, 365))
gpostneo <- eha::age.window(ginfant, c(g(30), g(365)))
print(postneo[1:5, ], row.names = FALSE) # print first five rows
print(gpostneo[1:5, ], row.names = FALSE) # first five rows after...
```

@@

### Transform back to the natural time scale

Use Equations \@ref(eq:haz) and \@ref(eq:cumhaz) to calculate the exogenous mortality distribution on 
the natural time scale.

```{r back}
H <- function(x, lambda = rate){
   lambda * g(x)
}

S <- function(x, lambda = rate){
   exp(-H(x, lambda))
}

getP <- function(lambda){
   days <- c(0, 0.25, 1:365)
   pd <- S(days, lambda)
   -diff(pd)
}
```

### Step 3: Estimate total and endogenous mortality

Now, we are primarily interested in the neonatal period, so let us zoom in on the 
neonatal age interval, Figure \@ref(fig:mag).

```{r mag, fig.cap = "Exogenous neonatal mortality, hazard and cumulative hazard functions."}
lam <- rate
oldpar <- par(mfrow = c(1, 2), cex.main = 0.9)
x <- seq(0, 30, length = 500)
plot(x, lam * dg(x) * 1000, type = "l", main = "Hazard function", col = "blue", 
     axes = FALSE, xlab = "Days", ylab = "per 1000")
axis(1, at = c(0, 15, 30))
axis(2, las = 1)
abline(h = 0, v = 0)
abline(v = 30, lty = 3, col = "darkgreen")
box()
plot(x, lam * g(x) * 1000, type = "l", main = "Cumulative hazard function", col = "blue",
     axes = FALSE, xlab = "Days", ylab = "per 1000")
axis(1, at = c(0, 15, 30))
axis(2, las = 1)
abline(h = 0, v = 0)
abline(v = 30, lty = 3, col = "darkgreen")
box()
par(oldpar)
```

@@

```{r hazards, fig.cap = "Cumulative hazard functions for neonatal mortality.", echo = TRUE}
neo <- age.window(infant, c(0, 30))
par(las = 1)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = neo)
hneo <- hazards(fit, cum = TRUE)
x <- c(0, hneo[[1]][, 1])
y <- c(0, 1000 * cumsum(hneo[[1]][, 2]))
plot(x, y, type = "l", xlab = "Day", ylab = "by 1000")
lines(x, 1000 * lam * g(x), col = "blue", lty = 2)
abline(h = 0, v = 0)
text(8, 15, "Total", col = "black")
text(8, 5, "Exogenous", col = "blue")
```

@@

```{r enof, fig.cap = "Cumulative hazard function for endogenous neonatal mortality.", fig.width = 7}
yen <- y - 1000 * lam * g(x)
par(las = 1)
plot(x, yen, type = "l", col = "red", xlab = "Day", ylab = "by 1000", ylim = c(0, 25.4))
abline(h = 0, v = 0)
lines(x, 1000 * lam * g(x), col = "blue", lty = 3)
lines(x, y, lty = 3)
text(8, 15, "Total", cex = 0.8)
text(15, 14, "Endogenous", col = "red")
text(8, 5, "Exogenous", col = "blue", cex = 0.8)
n <- length(yen)
axis(4, at = c(0, 1000 * lam * g(x[n]), yen[n], y[n]), las = 1, cex.axis = 0.7)
```

# The hazards approach in practice

@@

```{r lowume}
source("R/biometric.R")
ume <- infant##[infant$fodfrs %in% c(82750, 82780), ]
ume <- ume[ume$birthdate > "1920-12-31", ]
crudimr <- with(ume, sum(event) / sum(exit - enter)) * 365
crudprob <- round((1 - exp(-crudimr)) * 1000)
```

@@

```{r lowumee, fig.cap = "Västerbotten 1921--1950."}
##source("R/compare.R")
##xx <- bio(ume, timescale = "day", graph = TRUE)
##post <- age.window(ume, c(g(30), g(365)))
##rs.post <- risksets()
oldpar <- par(mfrow = c(1, 2))
biometric(ume, graph = TRUE, C = 365 / (log(366))^3)
##
par(oldpar)
```

There is a good model fit, and we see that the endogenous mortality clearly dominates 
the early days of neonatal mortality, and almost vanishes towardsthe start of the post-neonatal period.

@@

```{r bioume9120, fig.cap = "Västerbotten 1891--1920."}
##source("R/compare.R")
##xx <- bio(ume, timescale = "day", graph = TRUE)
##post <- age.window(ume, c(g(30), g(365)))
##rs.post <- risksets()
oldpar <- par(mfrow = c(1, 2))
biometric(ume, graph = TRUE, C = 365 / (log(366))^3)
##
par(oldpar)
```

The dominance of endogenous mortality in early life is still clear.

@@

```{r bioume6190, fig.cap = "Västerbotten 1861--1890."}
##source("R/compare.R")
##xx <- bio(ume, timescale = "day", graph = TRUE)
##post <- age.window(ume, c(g(30), g(365)))
##rs.post <- risksets()
oldpar <- par(mfrow = c(1, 2))
biometric(ume, graph = TRUE, C = 365 / (log(366))^3)
##
par(oldpar)
```

The conclusion here is almost the same as  for the later time periods, good model
fit and endogenous dominance in the very early days of life.

# Detecting under-registration

@@

This is an example of under-registration of early deaths.

```{r scahigh, fig.cap = "Scania 1710--1800."}
sca1 <- readRDS("Rdata/sca1.rds")
sca1$exit <- sca1$exit * 366 / (log(366))^3
sca1$enter <- sca1$enter * 366 / (log(366))^3
oldpar <- par(mfrow = c(1, 2))
biometric(sca1, graph = TRUE)
par(oldpar)
```

The estimation for the neonatal period breaks down totally, see Figure \@ref(fig:scahigh).
A sign of severe under-registration of neonatal deaths.

@@

Another example of severe under-registration, see Figure \@ref(fig:skum).

```{r skum, fig.cap = "Skellefteå 1821--1838."}
library(skum)
births <- obs[obs$starttyp == 2 &
                 obs$foddat > as.Date("1820-12-31") &
                 obs$foddat <= as.Date("1838-12-31"), ]
births$event <- births$sluttyp == 2
addup <- births$enter <= 0
births$enter[addup] <- births$enter[addup] + 0.0001
infa <- age.window(births, c(0, 1))
biometric(infa, timescale = "year", graph = TRUE)
```

# Known causes of death

In our data set U22003 we have some causes of death, distributed as in Table \@ref(tab:dcause).

```{r dcause, results='asis'}
deaths <- infant[infant$event == 1, ]
xx <- with(deaths, table(icd.chapter))
tt <- cbind(names(xx), as.vector(xx))
tbl(tt, rownames = FALSE, caption = "ICD-10 Category, frequencies.")
```

## Cumulative incidence analysis

Assuming that chapters XV, XVI, and XVII are endogenous and the rest is exogenous. 
Except for "missing value", which is a separate category, named "not available"
in Figure \@ref(fig:cuminc). This is a *competing risks analysis* [@ccr01].
The R package *cmprsk* [@cmprsk] was used to produce  Figure \@ref(fig:cuminc).

```{r cuminc, fig.cap = "Cumulative incidence functions, Västerbotten 1861-1950.", fig.height = 9}
par(mfrow = c(2, 2), lwd = 2)
library(survival)
infant$fstatus <- 0
endogen <- c("C", "D", "E", "G", "H", "O", "P", "Q")
endo <- (infant$icd.chapter %in% endogen) & (infant$event == 1)
exo <- (!(infant$icd.chapter %in% endogen)) & (infant$event == 1) &
   (!is.na(infant$icd.chapter))
unknown <- (is.na(infant$icd.chapter)) & (infant$event == 1)
infant$fstatus[endo] <- 1
infant$fstatus[exo] <- 2
infant$fstatus[unknown] <- 3
infant$fstatus <- factor(infant$fstatus, labels = c("censor", "endogenous", "exogenous", "not reported"))

>>>>

```{r logcube, fig.cap = "The log-cube transform of time in days versus the identity transform (dashed). Note that g(365) = 365.", fig.height = 7}
x <- seq(0, 365, length = 366)
plot(x, g(x), type = "l", xlab = "Days", ylab = "g(Days)", 
     col = "blue", axes = FALSE)
att <- c(0, 28, 183, 365)
axis(1, at = att, labels = att)
axis(2, las = 1, at = g(att), labels = c("0", "g(28)", "g(183)", "365"))
box()
lines(c(0, 365), c(0, 365), lty = 2)
abline(v = c(0, 365), h = c(0, 365))
abline(v = 28, h = g(28), col = "darkgreen", lty = 3)
```

@@

Assume that a cohort of infants is followed over time from birth to age one. There are no drop-outs 
(no right censoring except at age 365 days). The exact age in days at each observed death is noted and transformed by $g$, 
and due to imperfect time measuring there may be tied death ages. As an illustrative example, 
we use a data set from northern Sweden, covering the years 1861--1950.
```{r readdata}
infant <- readRDS("Rdata/infant.rds")
infant$period <- cut(infant$birthdate, 
                   breaks = as.Date(c("1860-12-31", "1890-12-31", 
                                      "1920-12-31", "1951-01-01")),
                   labels = c("1861-1890", "1891-1920", "1921-1950"))
infant <- infant[, c("birthdate", "sex", "enter", "exit", "event", "period", "icd.chapter")]
ginfant <- infant
ginfant$enter <- g(infant$enter)
ginfant$exit <- g(infant$exit)
```
For the first and last 30 years in this data set, we have the results shown in Figure \@ref(fig:first).

```{r first, fig.cap = "The periods 1861--1890 and 1921--1950 in Västerbotten, Bourgeois-Pichat method."}
source("R/drawBP.R")
oldpar <- par(mfrow = c(1,2))
##
r1 <- with(ginfant[ginfant$period == "1861-1890", ], eha::Surv(enter, exit, event))
r1 <- eha::risksets(r1, members = FALSE)
drawBP(r1, main = "1861-1890", ylim = c(0, 0.1))
##
##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
##r2 <- eha::risksets(r2, members = FALSE)
##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
##
r3 <- with(ginfant[ginfant$period == "1921-1950", ], eha::Surv(enter, exit, event))
r3 <- eha::risksets(r3, members = FALSE)
drawBP(r3, main = "1921-1950", ylim = c(0, 0.1))
par(oldpar)
```

The cumulative numbers of death are plotted against the
corresponding death ages on the $g$ time scale, and as is seen, the fit to linearity
after 28 days (68 on the $g$ scale) is excellent for both time periods.

@@

Instead of using the total number of births throughout in the denominator as in  the
Bourgeois-Pichat procedure, we suggest 
using the present risk-set size, that is, the the total number of infants still alive 
immediately prior to the death age under investigation. It is easily achieved by the use of 
the function `risksets` in the package `eha`. See Table \@ref(tab:risksets) for 
an illustration.
```{r risksets, echo = FALSE, results = 'asis'}
source("R/tbl.R")
lt1 <- with(r1, cbind(ginv(risktimes), round(risktimes, 1), size, n.events))
lt1 <- data.frame(lt1)
names(lt1) <- c("Day", "g(Day)", "Population", "Deaths")
tbl(lt1[1:7, ], caption = "Life table, 1861-1890.")
```
It starts with `r as.integer(lt1[1, 3])` live births, of which `r lt1[1, 4]` dies on the day of birth, 
on average at the age of six hours (a quarter of a day), and so on.

@@

```{r second, fig.cap = "The periods 1861--1890 and 1921--1950 in Västerbotten, hazards method."}
source("R/drawBB.R")
oldpar <- par(mfrow = c(1,2))
##
##r1 <- with(ginfant[ginfant$period == "1861-1890", ], eha::Surv(enter, exit, event))
##r1 <- eha::risksets(r1, members = FALSE)
drawBB(r1, main = "1861-1890", ylim = c(0, 0.115))
##
##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
##r2 <- eha::risksets(r2, members = FALSE)
##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
##
##r3 <- with(ginfant[ginfant$period == "1921-1950", ], eha::Surv(enter, exit, event))
##r3 <- eha::risksets(r3, members = FALSE)
drawBB(r3, main = "1921-1950", ylim = c(0, 0.115))
par(oldpar)
```

@@

```{r both, fig.cap = "Bourgeois-Pichat and 'hazards method' plots."}
oldpar = par(mfrow = c(1, 2), cex = 0.7, las = 1)
plot(r1$risktimes, cumsum(r1$n.events) / r1$size[1], type = "l", col = "blue",
     main = "1861-1890", ylim = c(0, 0.1), xlab = "g(Day)", ylab = "Cum. hazards")
lines(r1$risktimes, cumsum(r1$n.events / r1$size), type = "l", col = "red")
abline(h = 0, v = 0)
text(120, 0.08, "Hazard method", col = "red")
text(200, 0.02,"Bourgeois-Pichat", col = "blue")
##
plot(r3$risktimes, cumsum(r3$n.events) / r3$size[1], type = "l", col = "blue",
     main = "1921-1950", ylim = c(0, 0.1), xlab = "g(Day)", ylab = "Cum. hazards")
lines(r3$risktimes, cumsum(r3$n.events / r3$size), type = "l", col = "red")
abline(h = 0, v = 0)
text(120, 0.05, "Hazard method", col = "red")
text(200, 0.015,"Bourgeois-Pichat", col = "blue")
##
##
par(oldpar)
```

@@

We note that the Bourgeois-Pichat method requires that no right censorings (infants lost 
to follow-up before one year of age) are present, and no left truncations (infants under observation 
only from an age later than birth). The hazards approach, on the other hand, allows
left truncation and right censoring as long as they are *non-informative* in the usual sense. Often, though,
this not very important, since new-born babies tend to be geographically stationary
for their first year of life.

Throughout the rest of the paper, everything happens on the $g$ time scale.

## Post-neonatal mortality

 We concentrate on the *postneonatal* period, since that is the period where Bourgeois-Pichat claims that the *cumulative distribution function* (CDF) is *Uniform*, and we suggest that the CDF is *exponential*, that is, the *cumulative hazard function* is "uniform" (linear).

In order to see this, the data set from above is *left truncated* at age 28 days (at 68 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 30.

```{r postneo, fig.cap = "Exponential fits to postneonatal data, Västerbotten."}
oldpar <- par(mfrow = c(1, 2), cex = 0.7)
post <- age.window(ginfant, c(g(28), g(365)))
p1 <- post[post$period == "1861-1890", ]
p3 <- post[post$period == "1921-1950", ]
fit1 <- coxreg(Surv(exit - enter, event) ~ 1, data = p1)
wfit1 <- phreg(Surv(exit - enter, event) ~ 1, data = p1, shape = 1)
##check.dist(fit1, wfit1, main = "1861-1890", printLegend = FALSE)
plot(fit1, main = "1861-1890", printLegend = FALSE, xlab = "g(Day)", col = "blue")
whaz1 <- hazards(wfit1)
with(whaz1, lines(x, y, lty = 2, col = "red"))
legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
       col = c("blue", "red"), cex = 1)
##
fit3 <- coxreg(Surv(exit - enter, event) ~ 1, data = p3)
wfit3 <- phreg(Surv(exit - enter, event) ~ 1, data = p3, shape = 1)
whaz <- hazards(wfit3)
plot(fit3, main = "1921-1950", printLegend = FALSE, xlab = "g(Day)")
with(whaz, lines(x, y, lty = 2, col = "red"))
legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
       col = c("blue", "red"), cex = 1)
par(oldpar)
```

The *Exponential* fit is excellent, with a slight edge for the early period data, 
where post-neonatal mortality is high. However, one question pops up: If the exponential 
assumption is in fact true, will it also hold for subgroups of the data?

@@

```{r joined, fig.cap = "Exponential fit for two periods, Västerbotten."}
oldpar = par(cex = 0.7)
p13 <- post[post$period != "1891-1920", ]
p13$period <- factor(p13$period)
fit13 <- coxreg(Surv(exit - enter, event) ~ 1, data = p13)
wfit13 <- phreg(Surv(exit - enter, event) ~ 1, data = p13, shape = 1)
plot(fit13, main = "1861-1890 and 1921-1950", xlab = "g(Days)", printLegend = FALSE)
whaz13 <- hazards(wfit13)
with(whaz13, lines(x, y, lty = 2, col = "red"))
legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
       col = c("blue", "red"), cex = 1)
#
par(oldpar)
```

@@

## Exogeneous mortality

We now consider the whole infant age span (on the $g$ scale), and compare the uniform and exponential densities, especially of interest is the fraction of deaths that occur in the neonatal phase. @manfred04 argues that a less satisfactory property of the *B-P* model is that this fraction is constant, 
19.7 percent, independent of the overall level of infant mortality. This is not the case for the hazards model, see Figure \@ref(fig:exog).
For instance, for $p = 0.2$ the fraction in question is 21.5 percent.

@@

## Endogeneous mortality

In order to calculate the *endogeneous* infant mortality, a simple two-step procedure leads
to the goal.

@@

2.   On the full infant interval, estimate the total cumulative hazard rate $A(t)$ with the usual 
Nelson-Aalen estimator $\hat{A}(t)$ [@nelson72; @aalen78]. Then, subtract $H(t)$ from 
$\hat{A}(t)$ to get $\hat{E}(t)$.
\begin{equation}
\hat{E}(t) = \hat{A}(t) - \hat{\lambda} t, 0 < t < g(28).
\end{equation}
Note that subtraction and addition of competing risks are okay on the hazards scale, 
but not with probabilities, which is yet another argument in favor of the hazards approach.

@@

### Step 1: Estimate post-neonatal (exogenous) mortality

Post-neonatal mortality is the same as exogenous mortality in the sense that the
hazard functions are the same on the post-neonatal age interval. So the first step
involves data left truncated at age 28 days (or at `r round(g(28))`) on the $g$ scale.
We exemplify the numerical procedure with the aid of the **eha** [@eha; @ehar2].
The first lines of the data set are (Table \@ref(tab:firstlines))
```{r firstlines, results='asis'}

@@

```{r retrunc, echo = TRUE}
postneo <- eha::age.window(infant, c(28, 365))
gpostneo <- eha::age.window(ginfant, c(g(28), g(365)))
print(postneo[1:5, ], row.names = FALSE) # print first five rows
print(gpostneo[1:5, ], row.names = FALSE) # first five rows after...
```

@@

So we are done with the simple post-neonatal period.

### Step 2: Estimate total and endogenous mortality

Here we focus on neonatal period.

@@

```{r hazards, fig.cap = "Cumulative hazard functions for neonatal mortality.", echo = TRUE}
neo <- age.window(ginfant, c(0, g(28)))
par(las = 1)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = neo)
hneo <- hazards(fit, cum = TRUE)
x <- c(0, hneo[[1]][, 1])
y <- c(0, 1000 * cumsum(hneo[[1]][, 2]))
plot(x, y, type = "l", xlab = "Day", ylab = "by 1000")
lines(x, 1000 * rate * x, col = "blue", lty = 2)
abline(h = 0, v = 0)
text(8, 15, "Total", col = "black")
text(8, 5, "Exogenous", col = "blue")
```

@@

```{r enof, fig.cap = "Cumulative hazard function for endogenous neonatal mortality.", fig.width = 7}
yen <- y - 1000 * rate * x
par(las = 1)
plot(x, yen, type = "l", col = "red", xlab = "Day", ylab = "by 1000", ylim = c(0, 36))
abline(h = 0, v = 0)
lines(x, 1000 * rate * g(x), col = "blue", lty = 3)
lines(x, y, lty = 1)
text(8, 20, "Total", cex = 0.8)
text(15, 14, "Endogenous", col = "red")
text(8, 5, "Exogenous", col = "blue", cex = 0.8)
n <- length(yen)
axis(4, at = c(0, 1000 * rate * g(x[n]), yen[n], y[n]), las = 1, cex.axis = 0.7)
```

# Biometric analysis in practice

The original procedure of Bourgeos-Pichat is compared to the hazard based procedure for some typical
cases from the real world.

```{r plotfun}
pf <- function(x){
   ## x is a dataframe on g scale
   rs <- with(x, eha::risksets(Surv(enter, exit, event), members = FALSE))
   oldpar <- par(mfrow = c(1, 2))
   drawBB(rs, main = "Hazards method")
   drawBP(rs, main = "B-P method")
##
par(oldpar)
}
```

@@

```{r lowume}
##source("R/drawBP.R")
##source("R/drawBB.R")
ume <- infant##[infant$fodfrs %in% c(82750, 82780), ]
ume <- ume[ume$birthdate > "1920-12-31", ]
crudimr <- with(ume, sum(event) / sum(exit - enter)) * 365
crudprob <- round((1 - exp(-crudimr)) * 1000)
```

@@

```{r lowumee, fig.cap = "Västerbotten 1921--1950."}
##source("R/compare.R")
##xx <- bio(ume, timescale = "day", graph = TRUE)
##post <- age.window(ume, c(g(28), g(365)))
##rs.post <- risksets()
ume <- ginfant[ginfant$birthdate > "1920-12-31", ]
pf(ume)
```

There is a good hazards model fit, and we see that the endogenous mortality clearly dominates 
the early days of neonatal mortality, and almost vanishes towards the start of the post-neonatal period.

The B-P model fit is slightly worse, but not much to bother about.

@@

```{r bioume9120, fig.cap = "Västerbotten 1891--1920."}
##source("R/compare.R")
##xx <- bio(ume, timescale = "day", graph = TRUE)
##post <- age.window(ume, c(g(28), g(365)))
##rs.post <- risksets()
ume$exit <- g(ume$exit)
ume$enter <- g(ume$enter)
pf(ume)
```

The dominance of endogenous mortality in early life is still clear. The hazards fit is still slightly better.

@@

```{r bioume6190, fig.cap = "Västerbotten 1861--1890."}
##source("R/compare.R")
##xx <- bio(ume, timescale = "day", graph = TRUE)
##post <- age.window(ume, c(g(28), g(365)))
##rs.post <- risksets()
ume$exit <- g(ume$exit)
ume$enter <- g(ume$enter)
pf(ume)
```

The conclusion here is almost the same as  for the later time periods, good model
fit and endogenous dominance in the very early days of life. However, the B-P method seems to have a slight upper hand.

@@

```{r scahigh, fig.cap = "Scania 1710--1800."}
sca1 <- readRDS("Rdata/sca1.rds")
sca1 <- sca1[sca1$BirthDate < "1801-01-01", ]
sca1$exit <- g(sca1$exit)
sca1$enter <- g(sca1$enter)
rs <- with(sca1, risksets(Surv(enter, exit, event), members = FALSE))
oldpar <- par(mfrow = c(1, 2))
drawBB(rs, main = "Hazards method")
drawBP(rs, main = "B-P method")
par(oldpar)
```

Very bad fits in both cases.  <!--- Let us look at the estimated cumulative hazards, total, 
exogenous and endgenous, see Figure \@ref(fig:escu).

```{r escu, fig.cap = "Scania 1710-1800.", eval = FALSE}
#source("R/biometric.R")
#biometric(sca1, graph = TRUE)
post <- age.window(sca1, c(g(28), 365))
lambda <- with(post, sum(event) / sum(exit - enter))
x <- c(0, rs$risktimes)
y <- x * lambda
hazo <- with(rs, c(0, n.events / size))
hazo <- cumsum(hazo)
plot(x, hazo, type = "s", col = "blue")
lines(x, y)
z <- hazo - y
lines(x, z, col = "red")
```

--->

@@

An example of severe under-registration during the neonatal period, see Figure \@ref(fig:skum).
There is no room for endogeneous deaths at all, but otherwise a reasonably good fit with both 
methods.

```{r skum, fig.cap = "Skellefteå 1821--1838."}
library(skum)
births <- obs[obs$starttyp == 2 &
                 obs$foddat > as.Date("1820-12-31") &
                 obs$foddat <= as.Date("1838-12-31"), ]
births$event <- births$sluttyp == 2
births$enter <- 365 * births$enter
births$exit <- 365 * births$exit
addup <- births$exit <= 0
births$exit[addup] <- 0.25
infa <- age.window(births, c(0, 365))
infa$enter <- g(infa$enter)
infa$exit <- g(infa$exit)
pf(infa)
```
